---
title: "Lab 1: R Basics and indexing"
author: "Statistical Computing"
output: pdf_document
---

#change the false to TRUE and due in 1 week or 1.5 weeks and anything inside a $$ is LaTeX code

```{r, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, autodep=TRUE, cache.comments=TRUE)
```

Name: Leah Puglisi 
  Collaborated with:  Sam Gilletly
  
  This lab is to be done in class (completed outside of class if need be). You can collaborate with your classmates, but you must identify their names above, and you must submit **your own** lab as an knitted HTML file. To answer the questions, display the results and write your comments if asked.


```{r}
## For reproducibility --- don't change this!
set.seed(01182018)
```

**This lab's agenda**: manipulating data objects; using built-in functions, doing numerical calculations, and basic plots; reinforcing core probabilistic ideas; basic indexing, with a focus on matrices; some more basic plotting; vectorization; using `for()` loops.

The binomial distribution
===

The binomial distribution $\mathrm{Bin}(m,p)$ is defined by the number of successes in $m$ independent trials, each have probability $p$ of success. Think of flipping a coin $m$ times, where the coin is weighted to have probability $p$ of landing on heads.

The R function `rbinom()` generates random variables with a binomial distribution. E.g., 

```{r, eval=FALSE}
rbinom(n=20, size=10, prob=0.5)
```

produces 20 observations from $\mathrm{Bin}(10,0.5)$. If you change "eval=FALSE" to "eval=TRUE", it will run the code and display the numbers.

First half: some simple manipulations
===


- **1a.** Generate 200 random values from the $\mathrm{Bin}(10,0.5)$ distribution, and store them in a vector called `bin.draws.0.5`. Extract and display the first 10 elements. 

```{r, eval=TRUE}
bin_draws_0.5=rbinom(n=200, size=10, prob=0.5)
bin_draws_0.5[1:10]
```
\[a+b\] will display like LaTeX and in the middle of the page 

- **1b.** Compare the second element to the fifth, which is larger? A bit more tricky: print the indices of the elements of `bin.draws.0.5` that are equal to 5. How many such elements are there? **Challenge**: theoretically, how many such elements would you expect there to be?

```{r}
bin_draws_0.5[2]>bin_draws_0.5[5]
# this will give a false statement so then the second elemnt is not greater than the fifth
location=which(bin_draws_0.5==5)[] # put the putside[ ] to make sure it eill display as a vector and their locations
print(length(location)) #this will display the values 
# theoretically we could do a binomial probability count of x=5 and use the binomial equation so we should get areound the 46 number of 5's and about + or - 10 5's.  
```

- **1c.** Find the mean and standard deviation of `bin.draws.0.5`. Is the mean close what you'd expect? The standard deviation?

```{r}
xmean=mean(bin_draws_0.5)
xmean

xsd=sd(bin_draws_0.5)
xsd

#given the amount of 5's in the vector I am not too supprised by these numbers and the mean is np=5 and so we would expect the mean to be around 5 sd=np^2 which is the root(2.5) so yes those numbers make sense 
```
  
  - **1d.** Call `summary()` on `bin.draws.0.5` and describe the result.
```{r}
summary(bin_draws_0.5)
# this gives the 5-num-sum of the vector. the IQR is 8 and the mean and median are about the same with means we have a fairly normal and all the quantiles are the same unit distance from each other so this indicates a really normal plot.
```

- **1e.** Find the data type of the elements in `bin.draws.0.5` using `typeof()`. Then convert `bin.draws.0.5` to a vector of characters, storing the result as `bin.draws.0.5.char`, and use `typeof()` again to verify that you've done the conversion correctly. Call `summary()` on `bin.draws.0.5.char`. Is the result formatted differently from what you saw above? Why?

```{r}
typeof(bin_draws_0.5)

bin_draws_0.5.char=as.character(bin_draws_0.5)
summary(bin_draws_0.5.char)
#no they are they are not because the seed is the same and we are just creating integers rahter than a vector and then getting the legth of the vector counting all the characters, and then all others will be characters and wont have numerical values 

```

Some simple plots
===

- **2a.** The function `plot()` is a generic function in R for the visual display of data. The function `hist()` specifically produces a histogram display. Use `hist()` to produce a histogram of your random draws from the binomial distribution, stored in `bin.draws.0.5`. 

- **2b.** Call `tabulate()` on `bin.draws.0.5`. What is being shown? Does it roughly match the histogram you produced in the last question?

- **2c.** Call `plot()` on `bin.draws.0.5` to display your random values from the binomial distribution. Can you guess what the `plot()` function is doing here?

- **2d.** Call `plot()` with two arguments, the first being `1:200`, and the second being `bin.draws.0.5`. This creates a scatterplot of `bin.draws.0.5` (on the y-axis) versus the indices 1 through 200 (on the x-axis). Does this match your plot from the last question?

More binomials, more plots
===

- **3a.** Generate 200 binomials again, but change the probability of success to: 0.2, 0.3, 0.4, 0.6, 0.7, and 0.8, storing the results in vectors called `bin.draws.0.2`, `bin.draws.0.3`, `bin.draws.0.4.`, `bin.draws.0.6`, `bin.draws.0.7` and  `bin.draws.0.8`. For each, compute the mean and standard deviation.

- **3b.** We'd like to compare the properties of our vectors. Create a vector of length 7, whose entries are the means of the 7 vectors we've created, in order according to the success probabilities of their underlying binomial distributions (0.2 through 0.8).

- **3c.** Using the vectors from the last part, create the following scatterplots. Explain in words, for each, what's going on.
* The 7 means versus the 7 probabilities used to generate the draws.
* The standard deviations versus the probabilities.
* The standard deviations versus the means.

**Challenge**: for each plot, add a curve that corresponds to the relationships you'd expect to see in the theoretical population (i.e., with an infinite amount of draws, rather than just 200 draws).

Working with matrices
===

- **4a.** Create a matrix of dimension 200 x 7, called `bin.matrix`, whose columns contain the 7 vectors we've created, in order of the success probabilities of their underlying binomial distributions (0.2 through 0.8). Hint: use `cbind()`. 

- **4b.** Print the first three rows of `bin.matrix`. Print the element in the 77th row and 6th column. Compute the largest element in first column. Compute the largest element in all but the first column.

- **4c.** Calculate the column means of `bin.matrix` by using just a single function call.

- **4d.** Compare the means you computed in the last question to those you computed in Q3b, in two ways. First, using `==`, and second, using `identical()`. What do the two ways report? Are the results compatible? Explain.

- **4e.** Take the transpose of `bin.matrix` and then take row means. Are these the same as what you just computed? Should they be?
  
  