---
title: 'Lab 1: R Basics and indexing'
author: "Statistical Computing-Leah Puglisi"
output:
  html_document:
    df_print: paged
---

```{r, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, autodep=TRUE, cache.comments=TRUE)
```

Name: Leah Puglisi 
Collaborated with:  Sam Gilletly

```{r}
## For reproducibility --- don't change this!
set.seed(01182018)
```

**This lab's agenda**: manipulating data objects; using built-in functions, doing numerical calculations, and basic plots; reinforcing core probabilistic ideas; basic indexing, with a focus on matrices; some more basic plotting; vectorization; using `for()` loops.

The binomial distribution
===

The binomial distribution $\mathrm{Bin}(m,p)$ is defined by the number of successes in $m$ independent trials, each have probability $p$ of success. Think of flipping a coin $m$ times, where the coin is weighted to have probability $p$ of landing on heads.

The R function `rbinom()` generates random variables with a binomial distribution. E.g., 

```{r, eval=TRUE}
rbinom(n=20, size=10, prob=0.5)
```

Produces 20 observations from $\mathrm{Bin}(10,0.5)$.

First half: some simple manipulations
===


- **1a.** Generate 200 random values from the $\mathrm{Bin}(10,0.5)$ distribution, and store them in a vector called `bin.draws.0.5`. Extract and display the first 10 elements. 

```{r, eval=TRUE}
bin_draws_0.5=rbinom(n=200, size=10, prob=0.5)
bin_draws_0.5[1:10]
```

- **1b.** Compare the second element to the fifth, which is larger? A bit more tricky: print the indices of the elements of `bin.draws.0.5` that are equal to 5. How many such elements are there? **Challenge**: theoretically, how many such elements would you expect there to be?

```{r}
bin_draws_0.5[2]>bin_draws_0.5[5]
# This will give a FALSE statement so then the second element is not greater than the fifth. So element 5 is larger than element 2.
location=which(bin_draws_0.5==5)[] # put the putside[ ] to make sure it eill display as a vector and their locations
print(length(location)) #this will display the values 
# Theoretically we could do a binomial probability count of x=5 and use the binomial equation so we should get around 46 number of 5's and about + or - some 5's.  
```

- **1c.** Find the mean and standard deviation of `bin.draws.0.5`. Is the mean close what you'd expect? The standard deviation?

```{r}
xmean=mean(bin_draws_0.5)
xmean

xsd=sd(bin_draws_0.5)
xsd

#Given the amount of 5's in the vector I am not too supprised by these numbers and the mean is np=5 and so we would expect the mean to be around 5 sd=np^2 which is the root(2.5) so yes those numbers make sense. 
```
  
- **1d.** Call `summary()` on `bin.draws.0.5` and describe the result.
```{r}
summary(bin_draws_0.5)
# This gives the 5-num-sum of the vector. The IQR is 8 and the mean and median are about the same which means we have a fairly normal data distribution and all the quantiles are the same unit distance from each other so this indicates a really normal plot.
```

- **1e.** Find the data type of the elements in `bin.draws.0.5` using `typeof()`. Then convert `bin.draws.0.5` to a vector of characters, storing the result as `bin.draws.0.5.char`, and use `typeof()` again to verify that you've done the conversion correctly. Call `summary()` on `bin.draws.0.5.char`. Is the result formatted differently from what you saw above? Why?

```{r}
typeof(bin_draws_0.5)
bin_draws_0.5.char=as.character(bin_draws_0.5)
summary(bin_draws_0.5.char)
#No they are they are not because the seed is the same and we are just creating integers rahter than a vector and then getting the legth of the vector counting all the characters, and then all others will be characters and wont have numerical values. 

```

Some simple plots
===

- **2a.** The function `plot()` is a generic function in R for the visual display of data. The function `hist()` specifically produces a histogram display. Use `hist()` to produce a histogram of your random draws from the binomial distribution, stored in `bin.draws.0.5`. 

```{R}
hist(bin_draws_0.5)
# This is a right skewed distribution 
```



- **2b.** Call `tabulate()` on `bin.draws.0.5`. What is being shown? Does it roughly match the histogram you produced in the last question?

```{r}
tabulate(bin_draws_0.5)
#This is the number of occurences for the possible values of our sample from the binomial distirbution. In the histogram the values for 0,1,2 are combined into 1 column but you could still see a right skew in the numbers here. 
```


- **2c.** Call `plot()` on `bin.draws.0.5` to display your random values from the binomial distribution. Can you guess what the `plot()` function is doing here?

```{r}
plot(bin_draws_0.5)
#This is producing a 2-dimensional scatterplot with the numerical value of our sample values on the vertical axes and the index of our sample values on the horizontal axis. If not indicated, plot will always produce a scatterplot of some sort. 
```



- **2d.** Call `plot()` with two arguments, the first being `1:200`, and the second being `bin.draws.0.5`. This creates a scatterplot of `bin.draws.0.5` (on the y-axis) versus the indices 1 through 200 (on the x-axis). Does this match your plot from the last question?

```{r}
plot(1:200,bin_draws_0.5)
#Yes, this produces the same plot as above, it appears that when you give the plot function one argument it assumes the second value is the index, which in this case is a vector of integers from 1-200. 
```

More binomials, more plots
===

- **3a.** Generate 200 binomials again, but change the probability of success to: 0.2, 0.3, 0.4, 0.6, 0.7, and 0.8, storing the results in vectors called `bin.draws.0.2`, `bin.draws.0.3`, `bin.draws.0.4.`, `bin.draws.0.6`, `bin.draws.0.7` and  `bin.draws.0.8`. For each, compute the mean and standard deviation.

```{r}
bin_draws_0_2 = rbinom(200,10,0.2)
mean(bin_draws_0_2)
sd(bin_draws_0_2)

bin_draws_0_3 = rbinom(200,10,0.3)
mean(bin_draws_0_3)
sd(bin_draws_0_3)

bin_draws_0_4 = rbinom(200,10,0.4)
mean(bin_draws_0_4)
sd(bin_draws_0_4)

bin_draws_0_6 = rbinom(200,10,0.6)
mean(bin_draws_0_6)
sd(bin_draws_0_6)

bin_draws_0_7 = rbinom(200,10,0.7)
mean(bin_draws_0_7)
sd(bin_draws_0_7)

bin_draws_0_8 = rbinom(200,10,0.8)
mean(bin_draws_0_8)
sd(bin_draws_0_8)
```

- **3b.** We'd like to compare the properties of our vectors. Create a vector of length 7, whose entries are the means of the 7 vectors we've created, in order according to the success probabilities of their underlying binomial distributions (0.2 through 0.8).
    
```{r}
mean.vec = c(mean(bin_draws_0_2),mean(bin_draws_0_3),mean(bin_draws_0_4),mean(bin_draws_0_6),mean(bin_draws_0_7),mean(bin_draws_0_8))
```

- **3c.** Using the vectors from the last part, create the following scatterplots. Explain in words, for each, what's going on.
    * The 7 means versus the 7 probabilities used to generate the draws.
    * The standard deviations versus the probabilities.
    * The standard deviations versus the means.
    
    **Challenge**: for each plot, add a curve that corresponds to the relationships you'd expect to see in the theoretical population (i.e., with an infinite amount of draws, rather than just 200 draws).

```{r}
#Plot 1
prob.vec = c(0.2,0.3,0.4,0.6,0.7,0.8)
plot(mean.vec,prob.vec)
abline(lsfit(mean.vec,prob.vec),col="red")

#This plot produces a scatter plot of the 7 sample means verses the 7 sample binomial probabilities. This plot is very liner. As the probability of success for the sample increases, so does the mean from the sample. 
```


```{r}
#Plot 2
sd.vec = c(sd(bin_draws_0_2),sd(bin_draws_0_3),sd(bin_draws_0_4),sd(bin_draws_0_6),sd(bin_draws_0_7),sd(bin_draws_0_8))
plot(sd.vec,prob.vec)
abline(lsfit(sd.vec,prob.vec),col="red")

#This plot is like above but just with SD. The plot looks like a sideways porabola, so as the probability of success increases, the SD is either going up or down.  
```

```{r}
#Plot 3
plot(sd.vec,mean.vec)
abline(lsfit(sd.vec,mean.vec),col="red")
#This plot looks very much like the SD probability plot, which I expected as the mean and probability plot was so linear. 
```

Working with matrices
===

- **4a.** Create a matrix of dimension 200 x 7, called `bin.matrix`, whose columns contain the 7 vectors we've created, in order of the success probabilities of their underlying binomial distributions (0.2 through 0.8). Hint: use `cbind()`. 

```{r}
bin.matrix = cbind(bin_draws_0_2,bin_draws_0_3,bin_draws_0_4,bin_draws_0_6,bin_draws_0_7,bin_draws_0_8)
```

- **4b.** Print the first three rows of `bin.matrix`. Print the element in the 77th row and 6th column. Compute the largest element in first column. Compute the largest element in all but the first column.

```{r}
print(bin.matrix[1:3,])
print(bin.matrix[77,6])

max.val.col.1 = max(bin.matrix[,1])
max.val.col.1

max.val.other.columns = max(bin.matrix[,2:3])
max.val.other.columns
```

- **4c.** Calculate the column means of `bin.matrix` by using just a single function call.
```{r}
col.m = colMeans(bin.matrix)
col.m
```

- **4d.** Compare the means you computed in the last question to those you computed in Q3b, in two ways. First, using `==`, and second, using `identical()`. What do the two ways report? Are the results compatible? Explain.

```{r}
col.m == mean.vec

identical(col.m,mean.vec)

#There are not the same and no they are not compatible and one is a matrix and 1 is a vector.
```

- **4e.** Take the transpose of `bin.matrix` and then take row means. Are these the same as what you just computed? Should they be?

```{r}
trans.bin.matrix = t(bin.matrix)
print(rowMeans(trans.bin.matrix))
#Yes, they should be the same and they are the same as it is just the traspose of the matrix.
```


  
  