---
title: "Lab 9: Tidyverse I&II: Pipes and Dplyr"
author: "Statistical Computing"
---


```{r, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, autodep=TRUE, cache.comments=TRUE)
```

Name: Leah Puglisi 
Collaborated with:  Sam Gilletly 

This lab is to be done in class (completed outside of class if need be). You can collaborate with your classmates, but you must identify their names above, and you must submit **your own** lab as an knitted pdf file. To answer the questions, display the results and write your comments if asked.


**This week's agenda**: learning to master pipes, `dplyr`, grouping, spreading and gathering, and joins

```{r}
# Load the tidyverse!
library(tidyverse)
```

Pipes to base R
===

For each of the following code blocks, which are written with pipes, write equivalent code in base R (to do the same thing).

- **1a.** 

```{r}
# Pipes:
letters %>%
  toupper %>%
  paste(collapse="+") 

# Base R:

paste(toupper(letters), collapse = "+")
```

- **1b.** 

```{r}
# Pipes:
"     Ceci n'est pas une pipe     " %>% 
  gsub("une", "un", .) %>%
  trimws

# Base R:
trimws(gsub("une", "un", "Ceci n'est pas une pipe"))

```

- **1c.**

```{r}
# Pipes:
rnorm(1000) %>% 
  hist(breaks=30, main="N(0,1) draws", col="pink", prob=TRUE) 

# Base R:

hist(rnorm(1000), breaks = 30, main = "N(0,1) draws", col = "pink", prob = TRUE )
```

- **1d.** 

```{r}
# Pipes:
rnorm(1000) %>% 
  hist(breaks=30, plot=FALSE) %>%
  `[[`("density") %>%
  max
```

```{r}
# Base R:

max(hist(rnorm(1000), breaks = 30, plot=FALSE)$density)
```

Base R to pipes
===

For each of the following code blocks, which are written in base R, write equivalent code with pipes (to do the same thing).

- **2a.** Hint: you'll have to use the dot `.`, as seen above in Q1b, or in the lecture notes.

```{r}
# Base R:
paste("Your grade is", sample(c("A","B","C","D","R"), size=1))

# Pipes:
"Your grade is" %>% 
  paste(sample(c("A","B","C","D","R"),size=1))

sample(c("A","B","C","D","R"), size=1)  %>% 
  paste("Your grade is",.)
       
```

- **2b.** Hint: you can use the dot `.` again, in order to index `state.name` directly in the last pipe command.

```{r}
# Base R: 
state.name[which.max(state.x77[,"Illiteracy"])] 

# Pipes:
state.x77[,"Illiteracy"] %>%
  which.max() %>%
  state.name[.]
```

- **2c.** Hint: if `x` is a list of length 1, then `x[[1]]` is the same as `unlist(x)`.

```{r}
str.url = "http://www.stat.cmu.edu/~ryantibs/statcomp/data/trump.txt"

# Base R:
lines = readLines(str.url)
text = paste(lines, collapse=" ")
words = strsplit(text, split="[[:space:]]|[[:punct:]]")[[1]]
wordtab = table(words)
wordtab = sort(wordtab, decreasing=TRUE)
head(wordtab, 10)

# Pipes:
str.url %>%
readLines(.) %>%
paste(., collapse=" ") %>%
strsplit(., split="[[:space:]]|[[:punct:]]")  %>%
unlist(.) %>%
table(.) %>%
sort(., decreasing=TRUE)%>%
head(.,10)

```


Sprints data, revisited
===

Below we read in a data frame `sprint.w.df` containing the top women's times in the 100m sprint, as seen in previous labs. We also define a function `factor.to.numeric()` that was used in Lab 8, to convert the Wind column to numeric values. In what follows, use `dplyr` and pipes to answer the following questions on `sprint.w.df`.

```{r}
sprint.w.df = read.table(
  file="http://www.stat.cmu.edu/~ryantibs/statcomp/data/sprint.w.dat",
  sep="\t", header=TRUE, quote="", stringsAsFactors=TRUE)


factor.to.numeric = Vectorize(function(x) {
  x = strsplit(as.character(x), split = ",")[[1]]
  ifelse(length(x) > 1, 
         as.numeric(paste(x, collapse=".")), 
         as.numeric(x))
})
```

- **3a.** Convert the Wind column to numeric using `factor.to.numeric()`. Hint: use `mutate_at()`, and reassign `sprint.w.df` to be the output.

```{r}
sprint.w.df<-sprint.w.df %>%
  mutate_at(., "Wind",factor.to.numeric)
class(sprint.w.df$Wind)
```

- **3b.** Run a linear regression of the Time on Wind columns, but only using data where Wind values that are nonpositive, and report the coefficients. Hint: use `filter()`, and use the dot `.` to pipe into the `lm()` function appropriately.
```{r}
sprint.w.df %>%
  filter(., Wind <= 0) %>%
 lm(Wind~Time, data=.)

```
- **3c.** Plot the Time versus Wind columns, but only using data where Wind values that are nonpositive, and label the axes appropriately. Hint: recall that for a data frame, with columns `colX` and `colY`, you can use `plot(colY ~ colX, data=df)`, to plot `df$colY` versus `df$colX`.
```{r}
sprint.w.df %>%
  filter(., Wind <= 0) %>%
  plot(Time~Wind, data = ., main="Non-positive Wind Values v Time" )
```

- **3d.** Reorder the rows in terms of increasing Wind, and then display only the women who ran at most 10.7 seconds. Hint: do this with one single flow of pipe commands; use `arrange()`, `filter()`.
```{r}
sprint.w.df %>%
  arrange(., Wind) %>%
  filter(., Time <=10.7) %>%
  head(.)

```
- **3e.** Now reorder the rows in terms of increasing Time, and *then* increasing Wind, and again display only the women who ran at most 10.7 seconds, but only display the Time, Wind, Name, and Date columns. Hint: a single flow of pipe commands will do; note that `arrange()` can take multiple columns that you want to sort by, and the order you pass them specifies the priority.
```{r}
sprint.w.df %>%
  arrange(., Time, Wind) %>%
  filter(., Time <=10.7) %>%
  select(., Time, Wind, Name, Date) %>%
  head(.)

```

Practice with grouping
===

Below we read in a data frame `sprint.m.df` containing the top men's times in the 100m sprint, as seen in previous labs. In the following, unless stated otherwise, use pipes and `dplyr` verbs to solve each part as cleanly/succintly as you can.

```{r}
sprint.m.df = read.table(
  file="http://www.stat.cmu.edu/~ryantibs/statcomp/data/sprint.m.dat",
  sep="\t", header=TRUE, quote="", stringsAsFactors=TRUE)
```

- **4a.** Compute, for each country, the fastest time among athletes who come from that country, and display the first 10 results, ordered alphabetically by country. Also compute, for each city, the fastest time among athletes who ran in that city, and display the first 10 results, ordered alphabetically by city. Hint: `group_by()`, `summarise()`. 
```{r}
library(dplyr)

sprint.m.df %>%
  group_by(., Country) %>%
    summarise(.,min=min(Time)) %>%
head(., 10)

sprint.m.df %>%
  group_by(., City) %>%
    summarise(.,min=min(Time)) %>%
head(., 10)

```
- **4b.** With the most minor modification to your code possible, do the same computations as in the last part, but now display the first 10 results ordered by increasing time. Hint: `arrange()`.
```{r}
sprint.m.df %>%
group_by(., Country) %>%
  summarise(.,min=min(Time)) %>%
  arrange(., min) %>%
head(., 10)

sprint.m.df %>%
group_by(., City) %>%
  summarise(.,min=min(Time)) %>%
  arrange(., min) %>%
head(., 10)
```
- **4c.** Rewrite your solution in the last part using base R. Hint: `tapply()` gives probably the easiest route here. Note: your code here shouldn't be too much more complicated than your code in the last part.
```{r}

head(sort(tapply(sprint.m.df$Time, sprint.m.df$Country, min)), 10)

head(sort(tapply(sprint.m.df$Time, sprint.m.df$City, min)), 10)

```
- **4d.** Compute, for each country, the quadruple: (Name, City, Country, Time) corresponding to the athlete with the fastest time among athletes from that country. Display the first 10 results, ordered by increasing time. If there are ties, then show all the results that correspond to the fastest time. Repeat the same computation, but for the fastest time per city. Hint: `group_by()`, `filter()`, `select()`.
```{r}
sprint.m.df %>%
    group_by(.,Name, Country) %>%
        filter(., Time=min(Time)%>%
          select(., Name, City, Country, Time)%>%
head(., 10)

sprint.w.df %>%
    group_by(.,Name, Country) %>%
        filter(., Time=min(Time)%>%
          select(., Name, City, Country, Time)%>%
head(., 10)

```
- **4e.** Rewrite the rest of your solution in the last part using base R. You should end up with two data frames (per country, and per city) with the exact same structure as in the last part, and display the top 10 rows of each, ordered by increasing time. Hint: there are various routes to go; one strategy is to use `split()`, followed by `lapply()` with a custom function call, and then `rbind()` to get things in a data frame form. Note: your code here will probably be more complicated, or at least less intuitive, than your code in the last part.

```{r}

```
Practice spreading and gathering
===

In the following, use pipes and `dplyr` or `tidyr` verbs to solve each part as cleanly/succintly as you can. In some parts, it might make more sense to use direct indexing, and that's perfectly fine.

- **5a.** From `sprint.m.df`, define a reduced data frame `dat.reduced` as follows. For each athlete, and each city, compute the median of all times they recorded in this city. Your new data frame `dat.reduced` should have 1787 rows and 3 columns (Name, City, Time). Confirm that it has these dimensions, and display its first 10 entries. 
```{r}
library(dplyr)
library(tidyr)

dat.reduced <-sprint.m.df %>%
  select(., Name, City, Time) %>%
    group_by(., Name, City) %>%
      summarise(.,Time=median(Time)) 
head(dat.reduced, 10)
dim(dat.reduced)

```

- **5b.** The data frame `dat.reduced` is said to be in "long" format: it has observations on the rows, and variables (Name, City, Time) on the columns. Use `spread()` to convert this into "wide" format, and call the result `dat.wide`. Here the first column should be the athlete names, and the remaining columns should correspond to the cities. Apart from the first column, each entry gives the median time recorded by the athlete in this city. What are the dimensions of `dat.wide`, and do these make sense to you? 

```{r}
dat.wide<-
  dat.reduced %>%
    spread(., Time, City)
```

- **5c.** Not counting the names in the first column, how many non-`NA` values does `dat.wide` have? Does this make sense to you? It should. Reason how could you have guessed this number ahead of time, without even calling `spread()`, based only on `dat.reduced`?

```{r}

```

- **5d.** From `dat.wide`, look at the row for "Usain Bolt", and determine the city names that do not have `NA` values. These should be the cities in which he raced. Determine these cities directly from `dat.reduced`, and confirm that they match.

```{r}

```

- **5e.** Use `gather()` to convert `dat.wide` back into "long" format, and call the result `dat.long`. Remove rows that have `NA` values (hint: you can do this by setting `na.rm=TRUE` in the call to `gather()`), and order the rows alphabetically by athlete name. Once you've done this, `dat.wide` should have matching entries to `dat.reduced`; confirm that this is the case.

```{r}

```

Practice with joins
===

Below use data `sprint.w.df` and `sprint.m.df` again. In the following, use pipes and `dplyr` verbs to solve each part as cleanly/succintly as you can. Note: you'll receive warnings when you make joins about the conversion of factors to characters, and that's fine, don't worry about it.


- **6a.** Compute for each country, the triplet (Name, Country, Time) corresponding to the male athlete with the fastest time among athletes from that country, and breaking ties arbitrarily. Instead of displaying the results, save the resulting data frame as `dat.m`. Importantly, at the end of your flow of pipe commands used to define `dat.m`, make sure to call `ungroup()`. This will assure that `dat.m` has no groupings associated with it. Do the same for the women, and call the result `dat.w`. Report the dimensions of `dat.m` and `dat.w`, and check that they make sense to you.

```{r}
dat.m <-sprint.m.df %>%
  select(., Name, Country, Time) %>%
    group_by(.) %>%
      ungroup(.)%>%

dat.w<-sprint.w.df%>%
  select(., Name, Country, Time) %>%
    group_by(.) %>%
      ungroup(.)%>%
  
dim(dat.m)
dim(dat.w)
```

- **6b.** Perform an inner join, using `inner_join()`, of `dat.m` and `dat.w`, with the join done by the Country column. Call the resulting data frame `dat.ij`, and display its first 10 rows. How many rows does it have in total? Show how could you have arrived at this number ahead of time, from `dat.m$Country` and `dat.w$Country` (hint: `intersect()`). Count the number of `NA` values in `dat.ij`: this should be zero.

```{r}
dat.ij<-inner_join(dat.m, dat.w , by="Country")%>%
  head(., 10)
dim(dat.ij)


```

- **6c.** Perform a left join, using `left_join()`, of `dat.m` and `dat.w`, with the join again done by the Country column. Call the resulting data frame `dat.lj`, and display its first 10 rows. How many rows does it have in total? Explain why this makes sense. Count the number of `NA` values in `dat.lj`: this should be 50. Show how you could have arrived at this number from `dat.m$Country` and `dat.w$Country` (hint: `setdiff()`).

```{r}
dat.lj<-left_join(dat.m, dat.w , by="Country")%>%
  head(., 10)
dim(dat.lj)


```

- **6d.** Finally, perform an full join, using `full_join()`, of `dat.m` and `dat.w`, with the join again done by the Country column. Call the resulting data frame `dat.fj`. How many rows does it have in total? Show how you could have arrived at this number from `dat.m$Country` and `dat.w$Country` (hint: `union()`). Count the number of `NA` values in `dat.fj`: this should be 80. 

```{r}
dat.fj<-full_join(dat.m, dat.w , by="Country")%>%
  head(., 10)
dim(dat.fj)
```
