---
title: 'Lab 5: Functions'
author: "Leah Puglisi"
output:
  html_document:
    df_print: paged
---
```{r, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, autodep=TRUE, cache.comments=TRUE)
```

Name: Leah Puglisi
Collaborated with: Samuel Gilletly

This lab is to be done in class (completed outside of class if need be). You can collaborate with your classmates, but you must identify their names above, and you must submit **your own** lab as an knitted pdf file. To answer the questions, display the results and write your comments if asked.


Huber loss function
===

The Huber loss function (or just Huber function, for short) is defined as:
$$
\psi(x) = \begin{cases}
x^2 & \text{if $|x| \leq 1$} \\
2|x| - 1 & \text{if $|x| > 1$} 
\end{cases}
$$
This function is quadratic on the interval [-1,1], and linear outside of this interval. It transitions from quadratic to linear "smoothly". It is often used in place of the usual squared error loss for robust estimation. The sample average, $\bar{X}$---which given a sample $X_1,\ldots,X_n$ minimizes the squared error loss $\sum_{i=1}^n (X_i-m)^2$ over all choices of $m$---can be inaccurate as an estimate of $\mathbb{E}(X)$ if the distribution of $X$ is heavy-tailed. In such cases, minimizing Huber loss can give a better estimate. 

Some simple function tasks
===

- **1a.** Write a function `huber()` that takes as an input a number $x$, and returns the Huber value $\psi(x)$, as defined above. Hint: the body of a function is just a block of R code, e.g., in this code you can use `if()` and `else()` statements. Check that `huber(1)` returns 1, and `huber(4)` returns 7.

```{r}
huber = function(x){
  if(abs(x) <= 1){
    return(x^2)
  }else{
    return(2*abs(x)-1)
  }
}

huber(1)
huber(4)
```


- **1b.** The Huber function can be modified so that the transition from quadratic to linear happens at an arbitrary cutoff value $a$, as in:
$$
\psi_a(x) = \begin{cases}
x^2 & \text{if $|x| \leq a$} \\
2a|x| - a^2 & \text{if $|x| > a$} 
\end{cases}
$$
Starting with your solution code to the last question, update your `huber()` function so that it takes two arguments: $x$, a number at which to evaluate the loss, and $a$ a number representing the cutoff value. It should now return $\psi_a(x)$, as defined above. Check that `huber(3, 2)` returns 8, and `huber(3, 4)` returns 9.

```{r}
huber = function(x,a){
  if(abs(x)<=a){
    return(x^2)
  }else{
    return(2*a*abs(x)-a^2)
  }
}
huber(3,2)
huber(3,4)
```



- **1c.** Update your `huber()` function so that the default value of the cutoff $a$ is 1. Check that `huber(3)` returns 5.

```{r}
huber = function(x,a){
    if(abs(x)<=1){
    return(x^2)
  }else{
    return(2*abs(x)-1)
  }
}
huber(3)
```



- **1d.** Check that `huber(a=1, x=3)` returns 5. Check that `huber(1, 3)` returns 1. Explain why these are different.

```{r}
huber(a=1,x=3)
huber(1,3)
#These are different because if you don't specifiy which input corresponds to x and a it automatically assumes that the first is x and the second a, if you specify then it knows which is which. 
```



- **1e.** Vectorize your `huber()` function, so that the first input can actually be a vector of numbers, and what is returned is a vector whose elements give the Huber evaluated at each of these numbers. Hint: you might try using `ifelse()`, if you haven't already, to vectorize nicely. Check that `huber(x=1:6, a=3)` returns the vector of numbers (1, 4, 9, 15, 21, 27).

```{r}
huber = function(x,a){
  ifelse(abs(x)<=a,return(x^2),return(2*a*abs(x)-a^2))
}
huber(x=1:6,a=3)
```

- **Challenge.** Your instructor computed the Huber function values $\psi_a(x)$ over a bunch of different $x$ values, stored in `huber.vals` and `x.vals`, respectively. However, the cutoff $a$ was, let's say, lost. Using `huber.vals`, `x.vals`, and the definition of the Huber function, you should be able to figure out the cutoff value $a$, at least roughly. Estimate $a$ and explain how you got there. Hint: one way to estimate $a$ is to do so visually, using plotting tools; there are other ways too.

```{r}
x.vals = seq(0, 5, length=21)
huber.vals = c(0.0000, 0.0625, 0.2500, 0.5625, 1.0000, 1.5625, 2.2500,
               3.0625, 4.0000, 5.0625, 6.2500, 7.5625, 9.0000, 10.5000,
               12.0000, 13.5000, 15.0000, 16.5000, 18.0000, 19.5000, 
               21.0000)
```

Plotting practice, side effects 
===

- **2a.** Produce a plot of the basic `huber()` function ($a=1$) using the codes you wrote above. 

```{r}
plot(x= -10:10,huber(x=-10:10,a=1))
```



- **2b.** Modify the `huber()` function so that, as a side effect, it prints the string "Invented by the great Swiss statistician Peter Huber!" to the console. Hint: use `cat()`. Call your function on an input of your choosing, to demonstrate this side effect.

```{r}
huber = function(x,a){
  if(abs(x)<=a){
    cat("Invented by the great Swiss statistician Peter Huber!\n")
    return(x^2)
  }else{
    cat("Invented by the great Swiss statistician Peter Huber!\n")
    return(2*a*abs(x)-a^2)
  }
}
huber(1,1)
```


- **Challenge.** Further modify your `huber()` function so that, as another side effect, it produces a plot of Switzerland's national flag. Hint: look up this flag up on Google; it's pretty simple; and you should be able to recreate it with a few calls to `rect()`. Call your function on an input of your choosing, to demonstrate its side effects.

Exploring function environments
===

- **3a.** A modified version of the Huber function called `huber.mod()` is given below. You can see that we've defined the variable `x.squared` in the body of the function to be the square of the input argument `x`. In a separate line of code (outside of the function definition), define the variable `x.squared` to be equal to 999. Then call `huber.mod(x=3)`, and display the value of `x.squared`. What is its value? Is this affected by the function call `huber.mod(x=3)`? It shouldn't be! Reiterate this point with several more lines of code, in which you repeatedly define `x.squared` to be something different (even something nonnumeric, like a string), and then call `huber.mod(x=3)`, and demonstrate afterwards that the value of `x.squared` hasn't changed.

```{r}
huber.mod = function(x, a=1) {
  x.squared = x^2
  ifelse(abs(x) <= a, x.squared, 2*a*abs(x)-a^2)
}

x.squared = 999
huber.mod(x=3)
x.squared


x.squared = "Hello World!"
huber.mod(x=3)
x.squared

#`x.squared` retained its original value even though it was used within the function, indicating that when you declare somethingwithin the body of a function you don't change it's global declaration. 
```



- **3b.** Similar to the last question, define the variable `a` to be equal to -59.6, then call `huber.mod(x=3, a=2)`, and show that the value of `a` after this function call is unchanged. And repeat a few times with different assignments for the variable `a`, to reiterate this point.

```{r}
a = -59.6
huber.mod(x=3,a=2)
a

a = "Mandatory First Try"
huber.mod(x=3,a=2)
a

a = "Mandatory Second Try"
huber.mod(x=3,a=2)
a



```
- **3c.** The previous two questions showed you that a function's body has its own environment in which locally defined variables, like those defined in the body itself, or those defined through inputs to the function, take priority over those defined outside of the function. However, when a variable referred to the body of a function is *not defined in the local environment*, the default is to look for it in the global environment (outside of the function).

    Below is a "sloppy" implementation of the Huber function called `huber.sloppy()`, in which the cutoff `a` is not passed as an argument to the function. In a separate line of code (outside of the function definition), define `a` to be equal to 1.5 and then call `huber.sloppy(x=3)`. What is the output? Explain. Repeat this a few times, by defining `a` and then calling `huber.sloppy(x=3)`, to show that the value of `a` does indeed affect the function's ouptut as expected. **Challenge**: try setting `a` equal to a string and calling `huber.sloppy(x=3)`; can you explain what is happening?

```{r}
huber.sloppy = function(x) {
  ifelse(abs(x) <= a, x^2, 2*a*abs(x)-a^2)
}

a = 1.5
huber.sloppy(x=3)

a = 3
huber.sloppy(x=3)

a = "generic string"
huber.sloppy(x=3)
#When `a` changes it does change the output of the `huber.sloppy` function, however, when we change `a` to a string which would be incompatible with the function it seems to retain its previous numeric value.
```
 

- **3d.** At last, a difference between `=` and `<-`, explained! Some of you have asked about this. The equal sign `=` and assignment operator `<-` are often used interchangeably in R, and some people will often say that a choice between the two is mostly a matter of stylistic taste. This is not the full story. Indeed, `=` and `<-` behave very differently when used to set input arguments in a function call. As we showed above, setting, say, `a=5` as the input to `huber()` has no effect on the global assignment for `a`. However, replacing `a=5` with `a<-5` in the call to `huber()` is entirely different in terms of its effect on `a`. Demonstrate this, and explain what you are seeing in terms of global assignment.

```{r}
a = 1
cat("a's value:\n")
a
huber(x=3,a<-5)
cat("a's value:\n")
a
#Yes! This changed global `a` variables value! The global variable `a` no longer has a value `1` but now has a value `5` as defined within the call of the huber function.
```


- **3e.** The story now gets even more subtle. It turns out that the assignment operator `<-` allows us to define new global variables even when we are specifying inputs to a function. Pick a variable name that has not been defined yet in your workspace, say `b` (or something else, if this has already been used in your R Markdown document). Call `huber(x=3, b<-20)`, then display the value of `b`---this variable should now exist in the global enviroment, and it should be equal to 20! Also, can you explain the output of `huber(x=3, b<-20)`?

```{r}
huber(x=3,new_variable_name<-20)
new_variable_name
```
`huber(x=3,new_variable_name<-20)` results in an output of 9, indicating that even though the new variable we created is not called a it is being treated as `a` in our huber function. Additionally, this new variable now exists in the global environment. 


Shakespeare's complete works
===

Recall, as we saw in Week 3, that the complete works of [William Shakespeare](https://en.wikipedia.org/wiki/William_Shakespeare) are available freely from 
[Project Gutenberg](http://www.gutenberg.org). We've put this text file up at https://math.unm.edu/~lil/Stat590/labs/shakespeare.txt.

Getting lines of text play-by-play
===

- **4a.** Below is the `get.wordtab.from.url()` from lecture. Modify this function so that the string vectors `lines` and `words` are both included as named components in the returned list. For good practice, update the documentation in comments to reflect your changes. Then call this function on the URL for the Shakespeare's complete works (with the rest of the arguments at their default values) and save the result as `shakespeare.wordobj`. Using `head()`, display the first several elements of (definitely not all of!) the `lines`, `words`, and `wordtab` components of `shakespeare.wordobj`, just to check that the output makes sense to you.

```{r}
# get.wordtab.from.url: get a word table from text on the web
# Inputs:
# - str.url: string, specifying URL of a web page 
# - split: string, specifying what to split on. Default is the regex pattern
#   "[[:space:]]|[[:punct:]]"
# - tolower: Boolean, TRUE if words should be converted to lower case before
#   the word table is computed. Default is TRUE
# - keep.nums: Boolean, TRUE if words containing numbers should be kept in the
#   word table. Default is FALSE
# Output: list, containing lines, words, word table, and some basic summaries

get.wordtab.from.url = function(str.url, split="[[:space:]]|[[:punct:]]",
                                tolower=TRUE, keep.nums=FALSE) {
  lines = readLines(str.url)
  text = paste(lines, collapse=" ")
  words = strsplit(text, split=split)[[1]]
  words = words[words != ""]
    
  # Convert to lower case, if we're asked to
  if (tolower) words = tolower(words)
    
  # Get rid of words with numbers, if we're asked to
  if (!keep.nums) 
    words = grep("[0-9]", words, inv=TRUE, val=TRUE)
  
  # Compute the word table
  wordtab = table(words)
  
  return(list(wordtab=wordtab,
              number.unique.words=length(wordtab),
              number.total.words=sum(wordtab),
              longest.word=words[which.max(nchar(words))],
              #Now also returns the lines and words 
              lines = lines,
              words = words))
}



shakespeare.wordobj = get.wordtab.from.url(str.url = "https://math.unm.edu/~lil/Stat590/labs/shakespeare.txt")
summary(shakespeare.wordobj)
lapply(shakespeare.wordobj,head)
```
- **4b.** Define a function `get.wordtab.from.lines()` to have the same argument structure as `get.wordtab.from.url()`, except that the first argument of `get.wordtab.from.lines()` should be `lines`, a string vector passed by the user that contains lines of text to be processed. The body of `get.wordtab.from.lines()` should be the same as `get.wordtab.from.url()`, except that `lines` is passed and does not need to be computed using `readlines()`. The output of `get.wordtab.from.lines()` should be the same as `get.wordtab.from.url()`, except that `lines` does not need to be returned as a component. For good practice, incude documentation for your `get.wordtab.from.lines()` function in comments.

```{r}
get.word.tab.from.lines = function(lines, split="[[:space:]]|[[:punct:]]",
                                tolower=TRUE, keep.nums=FALSE){
  #lines = readLines(str.url)
  text = paste(lines, collapse=" ")
  words = strsplit(text, split=split)[[1]]
  words = words[words != ""]
    
  # Convert to lower case, if we're asked to
  if (tolower) words = tolower(words)
    
  # Get rid of words with numbers, if we're asked to
  if (!keep.nums) 
    words = grep("[0-9]", words, inv=TRUE, val=TRUE)
  
  # Compute the word table
  wordtab = table(words)
  
  return(list(wordtab=wordtab,
              number.unique.words=length(wordtab),
              number.total.words=sum(wordtab),
              longest.word=words[which.max(nchar(words))],
              #Now also returns words 
              words = words))
                                }


```




- **4c.** Look back at `get.wordtab.from.lines()` and `get.wordtab.from.url()`. Note that they overlap heavily, i.e., their bodies contain a lot of the same code. Redefine `get.wordtab.from.url()` so that it just calls `get.wordtab.from.lines()` in its body. Your new `get.wordtab.from.url()` function should have the same inputs as before, and produce the same output as before. So externally, nothing will have changed; we are just changing the internal structure of `get.wordtab.from.url()` to clean up our code base (specifically, to avoid code duplication in our case). This is an example of **code refactoring**. 

    Call your new `get.wordtab.from.url()` function on the URL for Shakespeare's complete works, saving the result as `shakespeare.wordobj2`. Compare some of the components of `shakespeare.wordobj2` to those of `shakespeare.wordobj` (which was computed using the old function definition) to check that your new implementation works as it should.

```{r}
get.wordtab.from.url = function(str.url, split="[[:space:]]|[[:punct:]]",
                                tolower=TRUE, keep.nums=FALSE) {
  lines = readLines(str.url)
  list = get.word.tab.from.lines(lines = lines)
  list$lines <- lines
  return(list)
}
shakespeare.word.obj2 = get.wordtab.from.url(str.url = "https://math.unm.edu/~lil/Stat590/labs/shakespeare.txt")
summary(shakespeare.word.obj2)
lapply(shakespeare.word.obj2,head)

```
- **Challenge.** Check using `all.equal()` whether `shakespeare.wordobj` and `shakespeare.wordobj2` are the same. Likely, this will not return `TRUE`. (If it does, then you've already solved this challenge question!) Modify your `get.wordtab.from.url()` function from the last question, so that it still calls `get.wordtab.from.lines()` to do the hard work, but produces an output exactly the same as the original `shakespeare.wordobj` object. Demonstrate your success by calling `all.equal()` once again